---
title: "Random weighted sample"
author: "Helen Bothwell"
date: "28 August 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code generates a random sample of points weighted by probability of habitat suitability. This code can also be used to generate a sample weighted by population density or other metric of interest. Code adapted from John Baumgartner (John.Baumgartner@mq.edu.au).

### Set working directory and load libraries
```{r}
setwd("C:/Path/to/my/data")
library(raster)
library(rgdal)
```

### Read in raster of probability of habitat suitability, population density, etc.
```{r}
r <- raster("Prob_Raster")
summary(r)
r
plot(r)
```

### Alternatively, you can generate a dummy raster. For example, a 10,000 x 100 matrix of random, uniform numbers between 0 and 1000.
```{r}
r <- raster(matrix(runif(1000000, 0, 1000), ncol=100))
```

### Change any NA values to zero. (This does not alter the original raster.)
```{r}
pr <- values(r)  #Extract values from all cells of r, and assign this vector of values to pr.
pr[is.na(pr)] <- 0  #Change NAs to zeroes.
```

### Take a random sample of n raster cells, weighted by probability of habitat suitability, population density, etc. (i.e. weighted by the values of the raster). seq_len(ncell(r)) produces a vector of the integers from 1 to the number of cells of r, where the probability weights of each of those numbers is equal to the values of pr.
```{r}
n <- 2500  #Define random, probability-weighted sample size.
cells <- sample(x=seq_len(ncell(r)), size=n, prob=pr)
```

### Create data.frame with a single column ('Habitat.Prob') containing just the probability of habitat suitability for selected cells. values(r)[cells] subsets the vector of probabilities, values(r), to just those values that correspond to the sampled cells.
```{r}
d <- data.frame(Habitat.Prob=values(r)[cells])
```

### Extract latitude and longitude of the chosen cells and return as a SpatialPoints object. The xyFromCell function identifies the cell-center coordinates for sampled cells of a raster object (in this case r, our raster of probability of habitat suitability). Setting spatial=TRUE returns the object as a SpatialPoints object, which facilitates writing out to a shapefile.
```{r}
coords <- xyFromCell(r, cells, spatial=TRUE)
```

### Combine coords with data to create SpatialPointsDataFrame object
Projection parameters reference: https://trac.osgeo.org/proj/wiki/GenParms
Find appropriate proj4 string: www.spatialreference.org
proj = Projection name (equidistant conic in this example)
lat_0 = Latitude of origin
lon_0 = Central meridian
lat_1 = Standard parallel 1
lat_2 = Standard parallel 2
x_0 = False easting
y_0 = False northing
ellps = Spheroid (GRS_1980)
no_defs = Don't use the /usr/share/proj/proj_def.dat defaults file (default of UTM Zone 10
NAD83 used if no_defs isn't specified)
```{r}
pts <- SpatialPointsDataFrame(coords, d, proj4string=CRS("+proj=eqdc +lat_0=39 +lon_0=- 107.7 +lat_1=33 +lat_2=45 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"))
plot(pts)
```

### Write to ESRI shapefile (requires rgdal package)
```{r}
writeOGR(pts, '.', 'Pts_prob_weighted', 'ESRI Shapefile')
```

