---
title: "PCA climate niche differentiation by pop"
author: "Helen Bothwell"
date: "3 November 2020"
output: html_document
---

This code plots population genetic groups in multivariate climate space using principal component analysis (PCA). The hypothesis that populations are significantly differentiated in climate space is then tested using perMANOVA.

### Call libraries and set working directory
```{r, message=FALSE}
library(psych)
library(vegan)
setwd("C:/Users/u1069501/Downloads/Modeling SDM paper/Analyses/PCA of genetic clusters in climate space")
```

### Read in data and define variables
```{r}
Data <- read.csv("NL_bioclim_top5_7popsFinal.csv")                           
head(Data)
climate <- Data[,c(3:7)]
summary(climate)
boxplot(climate,main="Boxplot of Bioclimatic Variables",xlab="Bioclimatic Variables")

population <- Data[c(8)]
site <- Data[c(2)]
Sample_ID <- Data[c(1)]
```

### Assess covariance and correlation between climate variables
```{r}
cov(climate)
cor_bioclim <- cor(climate)
write.csv(cor_bioclim,"cor_bioclim.csv")
```

### Standardize climate variables: z = (x - mean)/s.d.
```{r}
z.scores=vector()
for (i in 1:5){  #i in 1:nbioclim
x=climate[,i]
z.i=(x-mean(x))/sd(x)
z.scores=cbind(z.scores,z.i)
}
colnames(z.scores) <- c("TDiurnRng","TSeas","AveTWrmQtr","PSeas","PWrmQtr")
head(z.scores)
boxplot(z.scores,main="Boxplot of Normalized Bioclimatic Variables",xlab="Bioclimatic Variables",ylab="Z.Score")
```

### Run PCA, extract eigenvectors, eigenvalues, generate scree plot of PC contributions
```{r}
Fclimate <- principal(z.scores, nfactors=5, rotate="varimax", scores=T)
summary(Fclimate)
Fclimate
head(Fclimate$scores)  #Eigenvectors
Fclimate$values  #Eigenvalues
plot(Fclimate$values)
Eigenvalue <- Fclimate$values
plot(Eigenvalue,xlab="Principal Component",main="Scree Plot")
```

### Calculate % variance explained by PC axes
```{r}
percentagePC1=2.4910245/(2.4910245 + 1.0156676 + 0.9906276 + 0.4003561 + 0.1023242)
percentagePC1
percentagePC2=1.0156676/(2.4910245 + 1.0156676 + 0.9906276 + 0.4003561 + 0.1023242)
percentagePC2
percentagePC3=0.9906276/(2.4910245 + 1.0156676 + 0.9906276 + 0.4003561 + 0.1023242)
percentagePC3
percentagePC4=0.4003561/(2.4910245 + 1.0156676 + 0.9906276 + 0.4003561 + 0.1023242)
percentagePC4
percentagePC5=0.1023242/(2.4910245 + 1.0156676 + 0.9906276 + 0.4003561 + 0.1023242)
percentagePC5
PC1.PC2=percentagePC1+percentagePC2
PC1.PC2
PC1.PC2.PC3=percentagePC1+percentagePC2+percentagePC3
PC1.PC2.PC3
```

### Fit vector of sampling location centroids, plot in climate ordination space, and test site goodness of fit
```{r}
Fclimate.biplot <- envfit(Fclimate$scores, env=site) 
Fclimate.biplot
plot(Fclimate$scores)
plot(Fclimate.biplot)  #Add site ID labels
```

### Fit vector of sampling location centroids, plot in climate ordination space, and test population goodness of fit
```{r}
pop <- factor(population[,1])  
Fclimate.biplot <- envfit(Fclimate$scores,env=pop)
Fclimate.biplot
plot(Fclimate$scores)
plot(Fclimate.biplot)
```

### Run ANOVAs on the first PCs for site and population
```{r}
summary(aov(Fclimate$scores[,1] ~ factor(site[,1])))  
summary(aov(Fclimate$scores[,1] ~ factor(population[,1])))  
```

### 3D PCA Plot 
```{r}
library(rgl)
cols <- c('#5729CE','#CC33CC','#60DD49','#3F9B0B','#0074E7','aquamarine1','#FF9900')[as.numeric(factor(population[,1]),rotate="varimax")]
PCA_3D <- plot3d(Fclimate$scores,type='s',col=cols)
plot3d(Fclimate$scores,xlab="Precipitation Seasonality",ylab="Temperature Seasonality",zlab="Diurnal Temp. Range", type='s',col=cols)
```

### Save 3D image to .png file (file name must start with a letter - 3D_PCA gives error message)
```{r}
PCA_3D <- paste("pic",formatC(i,digits=1,flag="0"),".png",sep="")
rgl.snapshot(PCA_3D, fmt = "png", top = TRUE) #Same result as: rgl.snapshot(PCA_3D)
```

### PCA Plots w/ SE or SD ellipses 
```{r}
library(RColorBrewer)
display.brewer.all()
palette()  #Show what colors are being used by default
#http://mkweb.bcgsc.ca/colornames/
plot(Fclimate$scores,xlab="PC2 (20.3%)",ylab="PC1 (49.8%)",
     col=c('#5729CE','#CC33CC','#60DD49','#3F9B0B','#0074E7','aquamarine1','#FF9900')[as.numeric(factor(population[,1]),
     rotate="varimax")],pch=19)
```

### 1) Make points transparent, 2) Add standard error ellipses (smaller than SD below), 3)  Overlay climate vectors, 4) Add legend
```{r}
library(scales)
cols <- c('#5729CE','#CC33CC','#60DD49','#3F9B0B','#0074E7','aquamarine1','#FF9900')[as.numeric(factor(population[,1]), rotate="varimax")]
plot(Fclimate$scores,xlab="PC2 (20.3%)",ylab="PC1 (49.8%)",
     col=alpha(cols,0.4), pch=19, cex=1.5)  #alpha 0.4 = 40% opacity

ordiellipse(Fclimate$scores[population[,1] == '1',],factor(population[population[,1] == '1',1]),col='#5729CE')
ordiellipse(Fclimate$scores[population[,1] == '2',],factor(population[population[,1] == '2',1]),col='#CC00CC')
ordiellipse(Fclimate$scores[population[,1] == '3',],factor(population[population[,1] == '3',1]),col='#60DD49')
ordiellipse(Fclimate$scores[population[,1] == '4',],factor(population[population[,1] == '4',1]),col='#3F9B0B')
ordiellipse(Fclimate$scores[population[,1] == '5',],factor(population[population[,1] == '5',1]),col='#0074E7')
ordiellipse(Fclimate$scores[population[,1] == '6',],factor(population[population[,1] == '6',1]),col='aquamarine1')
ordiellipse(Fclimate$scores[population[,1] == '7',],factor(population[population[,1] == '7',1]),col='#FF9900')

fit <- envfit(Fclimate,climate,perm=999)
fit  #scores(fit, "vectors") #does same as "fit"
plot(fit, p.max = 0.05, col = "black",cex=1, pos=3)  #p.max displays only vectors significant at p < 0.05

legend("topleft", bty='n',
    legend=c("Arizona","New Mexico","Eastern Rocky Mnts","Western Colorado","Nevada & Central Utah","Idaho, W Wyoming, & N Utah","Montana & Alberta"),
    cex=0.8, col=alpha(c('#5729CE','#CC33CC','#60DD49','#3F9B0B','#0074E7','aquamarine1','#FF9900'),0.4), pch=19)
```

### Plot PCA w/ standard deviation ellipses of pts. w/ 95% condfidence limit (larger ellipses than SE)
```{r}
plot(Fclimate$scores,xlab="PC2 (20.3%)",ylab="PC1 (49.8%)",
     col=alpha(cols,0.4), pch=19, cex=1.5)
ordiellipse(Fclimate$scores[population[,1] == '1',],factor(population[population[,1] == '1',1]),kind="sd",conf=0.95,col='#5729CE')
ordiellipse(Fclimate$scores[population[,1] == '2',],factor(population[population[,1] == '2',1]),kind="sd",conf=0.95,col='#CC00CC')
ordiellipse(Fclimate$scores[population[,1] == '3',],factor(population[population[,1] == '3',1]),kind="sd",conf=0.95,col='#60DD49')
ordiellipse(Fclimate$scores[population[,1] == '4',],factor(population[population[,1] == '4',1]),kind="sd",conf=0.95,col='#3F9B0B')
ordiellipse(Fclimate$scores[population[,1] == '5',],factor(population[population[,1] == '5',1]),kind="sd",conf=0.95,col='#0074E7')
ordiellipse(Fclimate$scores[population[,1] == '6',],factor(population[population[,1] == '6',1]),kind="sd",conf=0.95,col='aquamarine1')
ordiellipse(Fclimate$scores[population[,1] == '7',],factor(population[population[,1] == '7',1]),kind="sd",conf=0.95,col='#FF9900')
```

### PCA spider plots
```{r}
plot(Fclimate$scores,main="PCA Spider Plot of Climate Niche by Population",xlab="PC2 (20.3%)",ylab="PC1 (49.8%)", col=alpha(cols,0.4), pch=19)
ordispider(Fclimate$scores[population[,1] == '1',],factor(population[population[,1] == '1',1]),col='#5729CE')
ordispider(Fclimate$scores[population[,1] == '2',],factor(population[population[,1] == '2',1]),col='#CC00CC')
ordispider(Fclimate$scores[population[,1] == '3',],factor(population[population[,1] == '3',1]),col='#60DD49')
ordispider(Fclimate$scores[population[,1] == '4',],factor(population[population[,1] == '4',1]),col='#3F9B0B')
ordispider(Fclimate$scores[population[,1] == '5',],factor(population[population[,1] == '5',1]),col='#0074E7')
ordispider(Fclimate$scores[population[,1] == '6',],factor(population[population[,1] == '6',1]),col='aquamarine1')
ordispider(Fclimate$scores[population[,1] == '7',],factor(population[population[,1] == '7',1]),col='#FF9900')
plot(fit, p.max = 0.05, col = "black",cex=1)
```

### Run perManova: Is the population binning process a good predictor of climate grouping? 
### Significant: Genetic population binning process is a good predictor of climate niche.
```{r}
pop <- factor(population[,1])  
adonis(climate ~ pop, permutations = 999, method="euc", strata=NULL) 
```

### Session Information
```{r}
options(width=100)
devtools::session_info()
```